---
title: "Fire Stations Layout"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true 
    theme: united
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---


\newpage

# Welcome

__Where there is smoke, there is fire__  

This is an R notebook written to support the paper _Bispo et al. (2023)  Using spatial point process models, clustering and space partitioning to reconfigure fire stations layout_. The main goal is to present the source code behind the study. 

This work is funded by national funds through the FCT - Fundação para a Ciência e a Tecnologia, I.P., under the scope of the project DSAIPA/DS/0088/2019 and research and development units UNIDEMI (project UIDB/00667/2020) and NOVAMATH (projects UIDB/00297/2020 and UIDP/00297/2020).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE,
	comment = " ",
	include = TRUE
)


library(ggplot2)
library(maps)
library(factoextra)
library(RgoogleMaps)
library(mapdata)
library(rworldmap)
library(maptools)
library(mapproj)
library(ggmap) 
library(rgdal)
library(viridis)
library(RColorBrewer)
library(tidygeocoder)
library(dplyr)
library(raster)
library(tmap)
library(rgeos)
library(sf)
library(spatstat)
library(geosphere)
library(readxl)
library(googleway)
library(deldir)
library(OpenStreetMap)
library(here)
library(rdist)
library(scapesClassification)
library(tidyr)
library(stars)
library(stringr)
```

# Functions 

This code chunk imports the functions defined in the R script _SourceFunctions.R_:

```{r source file}
source("Functions2FSLayout.R")
```

Session info:

```{r session info}
cat(paste("#",capture.output(sessionInfo()), "\n", collapse =""))
```

# Read files

## Shapefiles

```{r read shapefiles,include=FALSE,results='hide'}
unzip("shapefiles/dist/distritos.zip")
dist<-shapefile("distritos.shp")
dist$ID_0 <- as.factor(iconv(as.character(dist$ID_0), "UTF-8"))
dist$ISO <- as.factor(iconv(as.character(dist$ISO), "UTF-8"))
dist$NAME_0 <- as.factor(iconv(as.character(dist$NAME_0), "UTF-8"))
dist$ID_1 <- as.factor(iconv(as.character(dist$ID_1), "UTF-8"))
dist$NAME_1 <- as.factor(iconv(as.character(dist$NAME_1), "UTF-8"))
dist$HASC_1 <- as.factor(iconv(as.character(dist$HASC_1),  "UTF-8"))
dist$CCN_1<- as.factor(iconv(as.character(dist$CCN_1),  "UTF-8"))
dist$CCA_1 <- as.factor(iconv(as.character(dist$CCA_1), "UTF-8"))
dist$TYPE_1 <- as.factor(iconv(as.character(dist$TYPE_1), "UTF-8"))
dist$ENGTYPE_1 <- as.factor(iconv(as.character(dist$ENGTYPE_1), "UTF-8"))
dist$NL_NAME_1 <- as.factor(iconv(as.character(dist$NL_NAME_1), "UTF-8"))
dist$VARNAME_1 <- as.factor(iconv(as.character(dist$VARNAME_1), "UTF-8"))
dist<-dist[dist$NAME_1!="Madeira",]
dist<-dist[dist$NAME_1!="Açores",]
dist$NAME_1<-factor(str_to_title(dist$NAME_1))

unzip("shapefiles/freg/freguesias.zip")
freg<-sf::read_sf("freguesias.shp")
freg$Freguesia<-freg$Freguesia %>%
str_remove(pattern = "União das freguesias de ") %>%
str_remove(pattern = "União das freguesias da ") %>%
str_remove(pattern = "União de freguesias de ") %>%
str_remove(pattern = "União de freguesias da ") %>%
str_remove(pattern = "União de freguesias do ") %>%
str_remove(pattern = "União das freguesias das ") %>%
str_remove(pattern = "União das freguesias do ")
```

## Fire Stations Data

This code chunk reads the data files regarding fire stations (FS).

```{r fire stations data}
FireStations <- read_xlsx("FireStationsData.xlsx")
```

## Rural Fires Data

This code chunk reads the raw data files regarding rural fires between 2012 and 2020 and combine with information on fire stations.

```{r rural fires data}
rawfiles_rural<-list.files(pattern = "INFsumaria",recursive = FALSE)
rural<-list()
for (i in 1:length(rawfiles_rural)) {
  rural[[i]]<-read_xls(paste(rawfiles_rural[i]),skip = 6)[,c("Número","Distrito","Concelho","Freguesia","Latitude","Longitude")]
}
RuralData<-bind_rows(rural, .id = "column_label")
RuralData<-na.omit(merge(FireStations,RuralData,by = "Número"))
```

This code chunk change variables type character to factors.

```{r}
chr2fac<-colnames(select_if(RuralData, is.character))
RuralData[,chr2fac] <- lapply(RuralData[,chr2fac],factor)
```

The next chunk calculates the Euclidean Distances between rural fire events and FS.

```{r}
for (i in 1:nrow(RuralData)) {
RuralData$euc_dist[i] <- distm(c(RuralData$Longitude[i],RuralData$Latitude[i]),c(RuralData$LongitudeCB[i],RuralData$LatitudeCB[i]),fun=distHaversine)/1000
}
```

The next chunk adds the variable *type of fire* (rural/urban), removes uninformative variables, filters observations with distances under the 0.99 quantile and presents the summary of rural fires data.

```{r}
RuralData$Tipo<-factor(rep("rural",nrow(RuralData)))
RuralData$column_label<-NULL
RuralData$CodigoNatureza<-NULL
RuralData$EntidadeResponsavel<-NULL
RuralData = RuralData %>%  filter(RuralData$euc_dist<=quantile(RuralData$euc_dist,0.99))
RuralData$Distrito<-factor(str_to_title(RuralData$Distrito))
RuralData$Concelho<-factor(str_to_title(RuralData$Concelho))
```

There are `r nrow(RuralData)` rural fires.

## Urban Fires Data

```{r urban fires data}
UrbanData<-read_xlsx("UrbanFires.xlsx")
```

This code chunk change to factors variables type character.

```{r}
chr2fac<-colnames(select_if(UrbanData, is.character))
UrbanData[,chr2fac] <- lapply(UrbanData[,chr2fac],factor)
```

The next chunk calculates the Euclidean Distances between urban fire events and FS.

```{r}
for (i in 1:nrow(UrbanData)) {
UrbanData$euc_dist[i] <- distm(c(UrbanData$Longitude[i],UrbanData$Latitude[i]),c(UrbanData$LongitudeCB[i],UrbanData$LatitudeCB[i]),fun=distHaversine)/1000
}
```

The next chunk adds the variable *type of fire* (rural/urban), filters observations with distances above the 0.99 quantile and transform to lowercase letters administrative units names.

```{r}
UrbanData$Tipo<-factor(rep("urban",nrow(UrbanData)))
UrbanData <- UrbanData %>%  filter(UrbanData$euc_dist<=quantile(UrbanData$euc_dist,0.99))
UrbanData <- UrbanData[,c(names(RuralData))]
UrbanData$Distrito<-factor(str_to_title(UrbanData$Distrito))
UrbanData$Concelho<-factor(str_to_title(UrbanData$Concelho))
```

There are `r nrow(UrbanData)` urban fires.

## Population Data

```{r population data}
pop <- read_xlsx("covars.xlsx")
mapapop <- merge(freg,pop,by="Freguesia", all.x=T)
```

# Global Data

```{r global dataset}
Data <- rbind(RuralData,UrbanData)
Data$Distrito<-factor(str_to_title(Data$Distrito))
Data$Concelho<-factor(str_to_title(Data$Concelho))
Data$Freguesia<-factor(str_to_title(Data$Freguesia))
```

There are a total of `r nrow(Data)` fires.

## Cleaning

The next chunk limits observations to Portuguese borders.

```{r}
bound <- dist[dist$NAME_0=="Portugal",]
spdf <- SpatialPointsDataFrame(coords = Data[, c("Longitude", "Latitude")], data = Data, proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
Data <- Data %>%  filter(!is.na(over(spdf, as(bound, "SpatialPolygons"))))
```

The next chunks identifies the spatial points with coordinates in- and outside the respective given distric.

```{r}
namesd<-sort(unique(Data$Distrito))
listoutd<-list()
for (i in 1:length(namesd)) {
listoutd[[i]]=outlimitdist(paste(namesd[i]))
}
names(listoutd)<-namesd
dfs<-Filter(is.data.frame, listoutd)
df<-bind_rows(dfs)
rownames(df)<-NULL
```

The next chunk corrects incorrect district labeling.

```{r}
for (i in 1:length(dfs)) {
df=dfs[[i]]
spdf <- SpatialPointsDataFrame(coords = df[,c("Longitude", "Latitude")], data = df, proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
newloc<-as.character(dist$NAME_1[over(spdf, as(dist, "SpatialPolygons"))])
Data[Data$Longitude %in% df$Longitude & Data$Latitude %in% df$Latitude,"Distrito"]<-newloc
}
clistoutd<-list()
for (i in 1:length(namesd)) {
  clistoutd[[i]]=outlimitdist(paste(namesd[i]))
  }
```

Removing not availables entries:

```{r}
Data <- na.omit(Data)
```

## Split fire data by district

```{r data by district}
df.district <- split(Data, Data$Distrito)
```

## Spatial polygons dataframe by district

```{r spatial polygons}
df.region <- split(dist,dist$NAME_1)
```


# Application: Aveiro, Portugal

## Prepare data

-   Set region (window)

The next chunk retrieves the rectangular domain (space) under study.

```{r}
win <- setregion(df.region$Aveiro)
```

-   Dataset

The next chunk crops the observations within the area under study.

```{r}
Aveiro <- dborder(df.district$Aveiro)
Aveiro$Freguesia <- factor(Aveiro$Freguesia)
nc <- sum(table(unique(Aveiro$CodCB)))
```

## Point process by year

```{r}
df.y <- split(Aveiro, format(Aveiro$DataOcorrencia, "%Y"))
df <- lapply(df.y, "[", c("Longitude","Latitude"))
ppp.list <- lapply(df, pppfun, w=win[[1]])
ppp.list <- lapply(ppp.list, unique)
```

## Covariate population density

```{r}
rast<-st_rasterize(mapapop[mapapop$Distrito.x==toupper(names(df.district)[1]),] %>% 
  dplyr::select(PR, geometry))
r.pop <-raster(blur(as.im(rast)))
target<-ppp.list[[1]]
e <- extent(target$window$xrange[1],target$window$xrange[2],
            target$window$yrange[1],target$window$yrange[2])
SP.win <- as(e, "SpatialPolygons")
extent(r.pop)<-e
im<-as.im(r.pop)
poplist=list(pop=im)
```

The next chunk contains the code that generates the final hyperframe.

```{r}
H <- hyperframe(points=ppp.list,year=names(ppp.list),image=poplist$pop)
```


## Modelling

+ Population as an offset?

```{r,fig.height=4,fig.width=4}
rhos <- with(H, rhohat(points, image))
plot(rhos[[1]],main=" ")
```

+ Model

The next chunk adjusts an inhomogeneous Poisson process model with population density as an offset.

```{r}
model.list <- list(
fit1=mppm(points ~ x+y+offset(log(image)), H),
fit2=mppm(points ~ polynom(x,y,2)+offset(log(image)), H),
fit3=mppm(points ~ polynom(x,y,3)+offset(log(image)), H)
)
n.model<-as.numeric(which.min(lapply(model.list, AIC)))
fit<-model.list[[n.model]]
together<-subfits.new(fit)
```

## Simulation

Given the point process model fitted to data, generates random simulations of the model.

```{r}
set.seed(1)
nsim <- 50
rppp.list<- hyperframe(ppp=lapply(together, rmh, verbose=FALSE),iter=1)
for (j in 2:nsim) {
rppp<- hyperframe(ppp=lapply(together, rmh, verbose=FALSE),iter=j)
rppp.list<-rbind(rppp.list,rppp)
}
```

## Clustering

```{r}
df1 <- data.frame(long=rppp.list$ppp[[1]]$x, lat=rppp.list$ppp[[1]]$y)
set.seed(1)
km1 <- kmeans(df1,nc)
df1$clu <- km1$cluster
loc<-data.frame(km1$centers)
for (k in 2:length(rppp.list$ppp)) {
df2 <- data.frame(long=rppp.list$ppp[[k]]$x, lat=rppp.list$ppp[[k]]$y)
set.seed(1)
km2<- kmeans(df2,nc)
df2$clu <- km2$cluster
loc=cbind(loc, km2$centers[apply(cdist(km1$centers, km2$centers), 1, which.min),])
}
locmelt<-data.frame()
for (i in seq(1,(ncol(loc)-1),2)) {
   locmelt= rbind(locmelt,loc[,i:(i+1)])
}
```

The next chunk creates a point pattern to represent the locations in the two-dimensional plane and estimates the respective density.


```{r}
locppp<-ppp(locmelt$long,locmelt$lat,window=win[[1]])
bw<-bw.ppl(locppp)
dens<-density.ppp(locppp,sigma = bw,edge = TRUE,diggle = TRUE)
```

The code below finds the local maxima.

```{r,warning=FALSE}
w <- matrix(1,ncell(nc,d=dens),ncell(nc,d = dens))
r <- raster(dens)
localmax <- focal(r, w, fun = function(X) max(X, na.rm=TRUE), pad=FALSE, padValue=NA)
r2 <- r==localmax
maxXY <- xyFromCell(r2, Which(r2==1, cells=TRUE))
```

## Tesselation

```{r,fig.width=4,fig.height=5}
df <- data.frame(long=Aveiro$Longitude, lat=Aveiro$Latitude)
coordOPT <- data.frame(lon=maxXY[,1],lat=maxXY[,2])
coordCB <- data.frame(lon=unique(Aveiro)$LongitudeCB,lat=unique(Aveiro)$LatitudeCB)
d<-pointDistance(coordCB,coordOPT,lonlat=FALSE) 
dm <- as.matrix(d)
dmin <- apply(dm, 1, min, na.rm=TRUE)
wdmin <- apply(dm, 1, which.min)
v <- voronoipolygons(coordOPT, df.region$Aveiro)
proj4string(v) <- proj4string(df.region$Aveiro)
final <- gIntersection(df.region$Aveiro, v, byid=TRUE)
```


The code below the events to the new FS locations according to the new service areas by tesselation.

```{r}
d1<-pointDistance(df,coordOPT,lonlat=FALSE) 
dm1 <- as.matrix(d1)
dmin1 <- apply(dm1, 1, min, na.rm=TRUE)
wdmin1 <- apply(dm1, 1, which.min)
df$tile <- wdmin1
centers <- data.frame(longOPT=coordOPT[,1],latOPT=coordOPT[,2],tile=1:nrow(coordOPT))
alloc <- merge(df,centers,by="tile")
```

## Optimal sitting region

The code below defines an ellipse corresponding to a (1-alpha)\% confidence region for a bivariate normal distribution with mean _m_ and covariance _S_.


```{r}
alpha <-0.90
df.temp <- dplyr::filter(locmelt, long > -8.2 & lat > 40.9)
S <- cov(df.temp)
m <- colMeans(df.temp)
el <- ellipse::ellipse(S,centre = m,level=alpha)
```

## Distances

```{r}
distTP(reg = df.region$Aveiro,city = Aveiro,max = maxXY)
```

# Figures paper

```{r Figure 1}
par(mfrow=c(3,3))
par(oma=c(0,0,0,0))
par(mar=c(0,0,1.5,0))
for (i in 1:9) {
d<-density.ppp(ppp.list[[i]],sigma = 0.02,edge = TRUE,diggle = TRUE);plot(d,col = heat.colors(100,rev=T),main=paste(names(ppp.list)[i]),cex.main=1.5, ribbon=F,addcontour=F,clipwin=win[[1]],box=F);plot(win[[1]],add=T);points(ppp.list[[i]],cex=0.4,pch=16,col=2)
    }
```


```{r Figure 2(A)}
plot(rppp.list$ppp[[1]],main = "(A)",type="n"); points(ppp.list[[1]],cex=0.2,col=2)
``` 

```{r Figure 2(B)}
pred=predict.mppm(fit,ngrid = c(500,500))$trend[[1]]
plot(pred,col=heat.colors(300,rev=T),main="(B)",ribbon=F,box=F)
``` 

```{r Figure 2(C)}
plot(df1$long,df1$lat,pch=16,main="(C)",xlab=" ",ylab=" ",asp=0.9,axes=F,cex=0.5,col=terrain.colors(12)[11]);plot(win[[1]],add=T);for(i in 1:nc) {
    ConvexHull = chull(df1[df1$clu == i, ])
    polygon(df1[df1$clu == i, ][ConvexHull,],border=1)
};points(loc[,1],loc[,2],pch=4,cex=0.8,lwd=1.5)
``` 

```{r Figure 2(D)}
plot(dens,col = terrain.colors(300,rev=T),addcontour=F,box=F,ribbon=F,main="(D)");points(maxXY,pch=16,cex=0.5,col=1)
``` 

```{r Figure 2(E)}
plot(ppp.list[[1]],type="n",main="(E)");points(coordOPT$lon,coordOPT$lat,pch=16,col=1,cex=0.5);plot(final,add=T)

``` 

```{r Figure 2(F)}
plot(ppp.list[[1]],type="n",main="(F)");points(alloc$long,alloc$lat,type="p",cex=0.2,col=2,add=T);segments(alloc$long,alloc$lat,alloc$longOPT,alloc$latOPT,col=2);points(alloc$longOPT,alloc$latOPT,col=1,cex=0.5,pch=16)
```


```{r Figure 3}
plot(Aveiro$Longitude,Aveiro$Latitude,col=Aveiro$kmclus,xlab=" ",ylab=" ",cex.axis=1.2, cex.lab=1.2, asp=0.9,type="n")
lines(dist)
points(coordCB$lon,coordCB$lat,pch=1,cex=1.5)
points(coordOPT$lon,coordOPT$lat,pch=16,cex=1.5)
for (i in 1:nrow(coordCB)) {
 arrows(x0 = coordCB$lon[i],y0 = coordCB$lat[i],x1 = coordOPT$lon[wdmin[i]],y1 = coordOPT$lat[wdmin[i]],length = 0.1)
  }
bearing <- 1:360 * pi/180
cx <- (maxXY[21,1]+maxXY[16,1]+maxXY[19,1])/3 + 0.08*cos(bearing)
cy <- (maxXY[21,2]+maxXY[16,2]+maxXY[19,2])/3 + 0.08*sin(bearing)
circle <- cbind(cx, cy)
lines(circle, col='red', lwd=2)
text((maxXY[21,1]+maxXY[16,1]+maxXY[19,1])/3,(maxXY[21,2]+maxXY[16,2]+maxXY[19,2])/3,labels="(C)",cex=0.8,col=2,font=2)
cx <- maxXY[8,1] + 0.04*cos(bearing)
cy <- maxXY[8,2] + 0.04*sin(bearing)
circle <- cbind(cx, cy)
lines(circle, col='red', lwd=2)
text(maxXY[8,1] ,maxXY[8,2] ,labels="(B)",cex=0.8,pos=3,offset=0.4,col=2,font=2)
cx <- maxXY[4,1] + 0.03*cos(bearing)
cy <- maxXY[4,2] + 0.03*sin(bearing)
circle <- cbind(cx, cy)
lines(circle, col='red', lwd=2)
text(maxXY[4,1] ,maxXY[4,2] ,labels="(A)",cex=0.8,pos=2,offset=1.2,col=2,font=2)

```


The code below depicts the confidence region

```{r Figure 4}
par(mfrow=c(1,2))
par(oma=c(0,0,0,0))
par(mar=c(0,0,1,0))
plot(rppp.list$ppp[[1]], cex = 0.5,main = "(A)",type="n",cex.main=0.6)
points(df.temp,pch=16,cex=0.4)
for (i in c(0.90,0.99)) {
  lines(ellipse::ellipse(S,centre = m,level=i),type="l")
}
polygon(el[,1],el[,2], col = "blue",density = 20)
plot(rppp.list$ppp[[1]], cex = 0.5,main = "(B)",type="n",cex.main=0.6)
# points(df.temp,pch=16,cex=0.2)
for (i in c(0.90,0.99)) {
  lines(ellipse::ellipse(S,centre = m,level=i),type="l")
}
polygon(el[,1],el[,2], col = "blue",density = 20)
points(-8.242602,40.928327,pch=16,col=2)
points(m[1],m[2],col="blue",pch=16)
points(maxXY[6,1],maxXY[6,2],col="black",pch=16)
```
